---
title: "Summary Regression Models"
author: "Desiré De Waele"
date: "17 maart 2016"
output: html_document
---


- [Formulas](#Formulas)
- [Code for coefficients, prediction outcomes and residuals](#Code for coefficients, prediction outcomes and residuals)
- [Code for plotting a linear model fit and its residuals](#Code for plotting a linear model fit and its residuals)
- [Residual variation](#Residual variation)
- [Residual variation](#Residual variation)

##Used data summary 

The data used throughout this summary is the diamond dataset of the UsingR package. These plots show the relation between diamond weight (carat) and price in Dollars, the linear model fit and its residuals.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9, echo=FALSE}
library(ggplot2);library(UsingR);library(gridExtra);data(diamond)
y <- diamond$price; x <- diamond$carat; n <- length(y)

# Plotting data with their linea fit
g1 = ggplot(data.frame(x = x, y = y), aes(x = x, y = y))
g1 = g1 + geom_smooth(method = "lm", colour = "black")
g1 = g1 + geom_point(size = 3, colour = "black", alpha = 0.4)
g1 = g1 + geom_point(size = 2, colour = "red", alpha = 0.4)
g1 = g1 + xlab("Diamond weight in carats") + ylab("Singaporiang Dollars")

        
# Plotting the residuals of the data
g2 = ggplot(data.frame(x = x, y = resid(lm(y ~ x))), aes(x = x, y = y))
g2 = g2 + geom_hline(yintercept = 0, size = 1); 
g2 = g2 + geom_point(size = 3, colour = "black", alpha = 0.4)
g2 = g2 + geom_point(size = 2, colour = "red", alpha = 0.4)
g2 = g2 + xlab("Diamond weight in carats") + ylab("Residual Distances")

grid.arrange(g1, g2, ncol=2)
```

##Formulas

These are the formulas for variance, covariance and correlation.

$$ 
\begin{aligned} 
&S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar X)^2 = \frac{1}{n-1} \left( \sum_{i=1}^n X_i^2 - n \bar X ^ 2 \right)\\
&Cov(X, Y) = \frac{1}{n-1}\sum_{i=1}^n (X_i - \bar X) (Y_i - \bar Y) = \frac{1}{n-1}\left( \sum_{i=1}^n X_i Y_i - n \bar X \bar Y\right)\\
&Cor(X, Y) = \frac{Cov(X, Y)}{S_x S_y}
\end{aligned} 
$$

These are the formulas for linear modelfit, slope and intercept.

$$ 
\begin{aligned}
&Y_i = \beta_0 + \beta_1 X_i + \epsilon_i | \epsilon_i \sim N(0, \sigma^2)\\
&\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)}\\
&\hat \beta_0 = \bar Y - \hat \beta_1 \bar X\\
\end{aligned} 
$$

##Code for coefficients, prediction outcomes and residuals

Calculating coefficients of the linear fit by hand and with lm function.

```{r, message=FALSE, warning=FALSE}
y <- diamond$price; x <- diamond$carat; n <- length(y)
beta1 <- cor(y, x) * sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
fit <- lm(y ~ x)
rbind(c(beta0, beta1), coef(fit))
```

Predicting outcomes with a linear model fit by hand and with predict function

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
newx <- c(0.16, 0.27, 0.34)
byhand <- coef(fit)[1] + coef(fit)[2] * newx
byfunction <- predict(fit, newdata = data.frame(x = newx))
rbind(byhand, byfunction)
```

Calculating residuals by hand and with resid function.

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
byhand <- y - predict(fit)
byfunction <- resid(fit)
rbind(sort(byhand)[1:3], sort(byfunction)[1:3])
```

##Code for plotting a linear model fit and its residuals

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
# Generating the data
x = runif(100, -3, 3); y = x + sin(x) + rnorm(100, sd = .2); 

# Plotting data with their linea fit
g1 = ggplot(data.frame(x = x, y = y), aes(x = x, y = y))
g1 = g1 + geom_smooth(method = "lm", colour = "black")
g1 = g1 + geom_point(size = 3, colour = "black", alpha = 0.4)
g1 = g1 + geom_point(size = 2, colour = "red", alpha = 0.4)
        
# Plotting the residuals of the data
g2 = ggplot(data.frame(x = x, y = resid(lm(y ~ x))), aes(x = x, y = y))
g2 = g2 + geom_hline(yintercept = 0, size = 1); 
g2 = g2 + geom_point(size = 3, colour = "black", alpha = 0.4)
g2 = g2 + geom_point(size = 2, colour = "red", alpha = 0.4)
g2 = g2 + xlab("x") + ylab("Residual")

grid.arrange(g1, g2, ncol=2)
```

##Residual variation

Formula and code for calculating the residual variation, by hand and with formula.
$$ 
\begin{aligned}
&\hat \sigma^2 = \frac{\sum_{i=1}^n e_i^2}{n-2}\\
\end{aligned} 
$$

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
byhand <- sqrt(sum(resid(fit)^2) / (n - 2))
byformula <- summary(fit)$sigma
rbind(byhand, byformula)
```

The total variation sums the residual variation and the regression variation, r squared is the regression variation divided by the total variation.

$$ 
\begin{aligned}
&\sum_{i=1}^n (Y_i - \bar Y)^2 = \sum_{i=1}^n (Y_i - \hat Y_i)^2 + \sum_{i=1}^n (\hat Y_i - \bar Y)^2\\
&r^2 = \frac{\sum_{i=1}^n (\hat Y_i - \bar Y)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2} = Cor(X, Y)^2
\end{aligned} 
$$

##Standard error of slope and intercept
$$ 
\begin{aligned}
&\sigma_{\hat \beta_1}^2 = Var(\hat \beta_1) = \frac{\sigma^2}{\sum_{i=1}^n (X_i - \bar X)^2}\\
&\sigma_{\hat \beta_0}^2 = Var(\hat \beta_0) = \left(\frac{1}{n} + \frac{\bar X^2}{\sum_{i=1}^n (X_i - \bar X)^2 }\right)\sigma^2\\
\end{aligned} 
$$

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
beta1 <- cor(y, x) * sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
e <- y - beta0 - beta1 * x # Residuals
sigma <- sqrt(sum(e^2) / (n-2)) # Residual variation
ssx <- sum((x - mean(x))^2) # Denominator of coefficient standard errors

# Standard errors coefficients
seBeta0 <- (1 / n + mean(x) ^ 2 / ssx) ^ .5 * sigma 
seBeta1 <- sigma / sqrt(ssx)
# t-statistics when H0: beta mean = 0
tBeta0 <- beta0 / seBeta0; tBeta1 <- beta1 / seBeta1
# Calculating p-values
pBeta0 <- 2 * pt(abs(tBeta0), df = n - 2, lower.tail = FALSE)
pBeta1 <- 2 * pt(abs(tBeta1), df = n - 2, lower.tail = FALSE)

# Setting up a table
table <- rbind(c(beta0, seBeta0, tBeta0, pBeta0), c(beta1, seBeta1, tBeta1, pBeta1))
colnames(table) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")
rownames(table) <- c("(Intercept)", "x"); table

# The easy way
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x);
summary(fit)$coefficients
```
##Confidence interval for slope
```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
sumCoef <- summary(fit)$coefficients
# Calculating 95% interval for the price increase per 0.1 carat
(sumCoef[2,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[2, 2]) / 10
```


##Confidence and prediction intervals at given point

$$ 
\begin{aligned}
SE_{confidence} \ \text{at} \ x_0 &= \hat \sigma\sqrt{\frac{1}{n} + \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}\\
SE_{prediction} \ \text{at} \ x_0 &= \hat \sigma\sqrt{1 + \frac{1}{n} + \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}\\
\end{aligned} 
$$
Calculating the confidence interval (for the regression line) and prediction interval (for the data) at the mean of carat weight (x), by hand and with predict function with interval arguments.

```{r}
yhat <- fit$coef[1] + fit$coef[2] * mean(x)

confByHand <- yhat + c(-1, 1) * qt(.975, df = fit$df) * summary(fit)$sigma / sqrt(length(y))
confFormula <- predict(fit, newdata = data.frame(x = mean(x)), interval = ("confidence"))

predByHand <- yhat + c(-1, 1) * qt(.975, df = fit$df) * summary(fit)$sigma * sqrt(1 + 1/length(y))
predFormula <- predict(fit, newdata = data.frame(x = mean(x)), interval = ("prediction"))

table <- rbind(c(yhat, confByHand), confFormula, c(yhat, predByHand), predFormula)
rownames(table) <- c("Confidence by hand", "Confidence by formula", 
                     "Prediction by hand", "Prediction by formula")
colnames(table) <- c("Estimate", "Lower limit", "Upper limit"); table

```

Example plot with confidence and prediction intervals, to help interpret them.

```{r, fig.height=4, fig.width=7, echo=FALSE}
newx = data.frame(x = seq(min(x), max(x), length = 100))
p1 = data.frame(predict(fit, newdata= newx,interval = ("confidence")))
p2 = data.frame(predict(fit, newdata = newx,interval = ("prediction")))
p1$interval = "confidence"
p2$interval = "prediction"
p1$x = newx$x
p2$x = newx$x
dat = rbind(p1, p2)
names(dat)[1] = "y"

g = ggplot(dat, aes(x = x, y = y))
g = g + geom_ribbon(aes(ymin = lwr, ymax = upr, fill = interval), alpha = 0.2) 
g = g + geom_line()
g = g + geom_point(data = data.frame(x = x, y=y), aes(x = x, y = y), 
                   colour = "black", alpha = 0.4, size = 3)
g = g + geom_point(data = data.frame(x = x, y=y), aes(x = x, y = y), 
                   colour = "darkblue", alpha = 0.4, size = 2)
g
```