---
title: "Summary Regression Models Course"
author: "Desiré De Waele"
date: "17 maart 2016"
output:
  pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 2
---
#Plotting Linear Model and Residuals

Example of linear regression fit, and the residuals. Code is hidden for now.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9, echo=FALSE}
library(ggplot2); library(UsingR); library(gridExtra)
x = runif(100, -3, 3); y = x + sin(x) + rnorm(100, sd = .2); 

g1 = ggplot(data.frame(x = x, y = y), aes(x = x, y = y)) + theme_bw()
g1 = g1 + geom_smooth(method = "lm", colour = "black")
g1 = g1 + geom_point(size = 3, colour = "black", alpha = 0.4)
g1 = g1 + geom_point(size = 2, colour = "red", alpha = 0.4)
        
g2 = ggplot(data.frame(x = x, y = resid(lm(y ~ x))), aes(x = x, y = y)) + theme_bw()
g2 = g2 + geom_hline(yintercept = 0, size = 1); 
g2 = g2 + geom_point(size = 3, colour = "black", alpha = 0.4)
g2 = g2 + geom_point(size = 2, colour = "red", alpha = 0.4)
g2 = g2 + labs(y = "Residual")

grid.arrange(g1, g2, ncol=2)
```

The same code is used on the diamond dataset of the UsingR package, the data used throughout this summary. The plots below show the relation between carats and price, the linear model fit and its residuals.
```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
library(ggplot2); library(UsingR); library(gridExtra); data(diamond)
y <- diamond$price; x <- diamond$carat; n <- length(y)

g1 = ggplot(data.frame(x = x, y = y), aes(x = x, y = y)) + theme_bw()
g1 = g1 + geom_smooth(method = "lm", colour = "black")
g1 = g1 + geom_point(size = 3, colour = "black", alpha = 0.4)
g1 = g1 + geom_point(size = 2, colour = "red", alpha = 0.4)
g1 = g1 + labs(x = "Diamond Weight in Carats", y = "Singaporiang Dollars")

g2 = ggplot(data.frame(x = x, y = resid(lm(y ~ x))), aes(x = x, y = y)) + theme_bw()
g2 = g2 + geom_hline(yintercept = 0, size = 1); 
g2 = g2 + geom_point(size = 3, colour = "black", alpha = 0.4)
g2 = g2 + geom_point(size = 2, colour = "red", alpha = 0.4)
g2 = g2 + labs(x = "Diamond weight in carats", y = "Residual Distances")
grid.arrange(g1, g2, ncol=2)
```

#Linear Model
##Coefficients

Formulas for variance, covariance and correlation.

$$ 
\begin{aligned} 
&S^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i - \bar X)^2 = \frac{1}{n-1} \left( \sum_{i=1}^n X_i^2 - n \bar X ^ 2 \right)\\
&Cov(X, Y) = \frac{1}{n-1}\sum_{i=1}^n (X_i - \bar X) (Y_i - \bar Y) = \frac{1}{n-1}\left( \sum_{i=1}^n X_i Y_i - n \bar X \bar Y\right)\\
&Cor(X, Y) = \frac{Cov(X, Y)}{S_x S_y}
\end{aligned} 
$$

Formulas for linear model fit, slope and intercept.

$$ 
\begin{aligned}
&Y_i = \beta_0 + \beta_1 X_i + \epsilon_i | \epsilon_i \sim N(0, \sigma^2)\\
&\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)}\\
&\hat \beta_0 = \bar Y - \hat \beta_1 \bar X\\
\end{aligned} 
$$

Calculating coefficients of the linear fit by hand and with lm function.

```{r, message=FALSE, warning=FALSE}
y <- diamond$price; x <- diamond$carat; n <- length(y)
beta1 <- cor(y, x) * sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
fit <- lm(y ~ x)
rbind(c(beta0, beta1), coef(fit))
```
##Outcome Estimates

By hand and with predict function.

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
newx <- c(0.16, 0.27, 0.34)
byhand <- coef(fit)[1] + coef(fit)[2] * newx
byfunction <- predict(fit, newdata = data.frame(x = newx))
rbind(byhand, byfunction)
```

#Residuals

##Calculation

By hand and with resid function.

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
byhand <- y - predict(fit)
byfunction <- resid(fit)
rbind(sort(byhand)[1:3], sort(byfunction)[1:3])
```


##Residual Variation

Formula for residual variation.
$$ 
\begin{aligned}
&\hat \sigma^2 = \frac{\sum_{i=1}^n e_i^2}{n-2}\\
\end{aligned} 
$$

Calculating the residual variation with formula, or by retrieval of lm function.

```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
byhand <- sqrt(sum(resid(fit)^2) / (n - 2))
byformula <- summary(fit)$sigma
rbind(byhand, byformula)
```

## Regression Variation
The total variation sums the residual variation and the regression variation. The latter is the variance explained by the linear model. R squared is the regression variation divided by the total variation, and estimates the power of the linear model.

$$ 
\begin{aligned}
&\sum_{i=1}^n (Y_i - \bar Y)^2 = \sum_{i=1}^n (Y_i - \hat Y_i)^2 + \sum_{i=1}^n (\hat Y_i - \bar Y)^2\\
&r^2 = \frac{\sum_{i=1}^n (\hat Y_i - \bar Y)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2} = Cor(X, Y)^2
\end{aligned} 
$$

#Intervals

##Confidence Intervals for Coefficients

Formulas for standard errors of both the slope and the intercept.
$$ 
\begin{aligned}
&\sigma_{\hat \beta_1}^2 = Var(\hat \beta_1) = \frac{\sigma^2}{\sum_{i=1}^n (X_i - \bar X)^2}\\
&\sigma_{\hat \beta_0}^2 = Var(\hat \beta_0) = \left(\frac{1}{n} + \frac{\bar X^2}{\sum_{i=1}^n (X_i - \bar X)^2 }\right)\sigma^2\\
\end{aligned} 
$$

Setting up a coefficient table by hand...
```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
beta1 <- cor(y, x) * sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
e <- y - beta0 - beta1 * x # Residuals
sigma <- sqrt(sum(e^2) / (n-2)) # Residual variation
ssx <- sum((x - mean(x))^2) # Denominator of coefficient standard errors

# Standard errors coefficients
seBeta0 <- (1 / n + mean(x) ^ 2 / ssx) ^ .5 * sigma 
seBeta1 <- sigma / sqrt(ssx)
# t-statistics when H0: beta mean = 0
tBeta0 <- beta0 / seBeta0; tBeta1 <- beta1 / seBeta1
# Calculating p-values
pBeta0 <- 2 * pt(abs(tBeta0), df = n - 2, lower.tail = FALSE)
pBeta1 <- 2 * pt(abs(tBeta1), df = n - 2, lower.tail = FALSE)

# Setting up a table
table <- rbind(c(beta0, seBeta0, tBeta0, pBeta0), c(beta1, seBeta1, tBeta1, pBeta1))
colnames(table) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")
rownames(table) <- c("(Intercept)", "x"); table
```
...or by retrieval of the lm function.
```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x);
summary(fit)$coefficients
```

###Calculation of slope confidence interval
```{r}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
sumCoef <- summary(fit)$coefficients
# Calculating 95% interval for the price increase per 0.1 carat
(sumCoef[2,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[2, 2]) / 10
```

##Confidence Interval for Linear Model

Formula for standard error of the confidence interval at a given point $x_0$.

$$ 
\begin{aligned}
SE_{confidence} \ \text{at} \ x_0 &= \hat \sigma\sqrt{\frac{1}{n} + \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}\\
\end{aligned} 
$$
Calculating the confidence interval at the mean of x, by hand and with predict function.

```{r}
yhat <- fit$coef[1] + fit$coef[2] * mean(x)
confByHand <- yhat + c(-1, 1) * qt(.975, df = fit$df) * summary(fit)$sigma / sqrt(length(y))
confFormula <- predict(fit, newdata = data.frame(x = mean(x)), interval = ("confidence"))

table <- rbind(c(yhat, confByHand), confFormula)
rownames(table) <- c("Confidence by hand", "Confidence by formula")
colnames(table) <- c("Estimate", "Lower limit", "Upper limit"); table
```

##Prediction Interval for Outcome Estimates

Formula for standard error of the prediction interval at a given point $x_0$.

$$ 
\begin{aligned}
SE_{prediction} \ \text{at} \ x_0 &= \hat \sigma\sqrt{1 + \frac{1}{n} + \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}\\
\end{aligned} 
$$
Calculating the prediction interval at the mean of x, by hand and with predict function.

```{r}
yhat <- fit$coef[1] + fit$coef[2] * mean(x)
predByHand <- yhat + c(-1, 1) * qt(.975, df = fit$df) * summary(fit)$sigma * sqrt(1 + 1/length(y))
predFormula <- predict(fit, newdata = data.frame(x = mean(x)), interval = ("prediction"))

table <- rbind(c(yhat, predByHand), predFormula)
rownames(table) <- c("Prediction by hand", "Prediction by formula")
colnames(table) <- c("Estimate", "Lower limit", "Upper limit"); table
```

Example plot with confidence and prediction intervals, to help interpret them.

```{r, fig.height=3, echo=FALSE}
newx = data.frame(x = seq(min(x), max(x), length = 100))
p1 = data.frame(predict(fit, newdata= newx,interval = ("confidence")))
p2 = data.frame(predict(fit, newdata = newx,interval = ("prediction")))
p1$interval = "confidence"
p2$interval = "prediction"
p1$x = newx$x
p2$x = newx$x
dat = rbind(p1, p2)
names(dat)[1] = "y"

g = ggplot(dat, aes(x = x, y = y))
g = g + geom_ribbon(aes(ymin = lwr, ymax = upr, fill = interval), alpha = 0.2) + theme_bw()
g = g + geom_line()
g = g + geom_point(data = data.frame(x = x, y=y), aes(x = x, y = y), 
                   colour = "black", alpha = 0.4, size = 2)
g = g + geom_point(data = data.frame(x = x, y=y), aes(x = x, y = y), 
                   colour = "grey", alpha = 0.4, size = 1.5)
g
```